A free/open-source Kazakh-Tatar machine translation system

### 2013-04-19 paper/paper.tex
мен үйге қайтуым керек <-> миңа өйгә кайтырга кирәк; add a note

Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are different.  Specifically, in both languages, the person agreement takes the form of a person copula suffix, although in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in Tatar this suffix intervenes between the two pieces of the ``compound'' tense morpheme (e.g., баргансың икән ``I guess you went'').

@@

%\begin{exe}
%	\setstretch{0.8}
%	\ex
%	\begin{xlist}
%		\ex Мен үйге баруым керек
%		\gll мен үй-GA бар-у-Iм керек \\
%		I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need \\
%		\glt \eng{I need to go home}
%
%		\ex Мин өйгә барырга кирәк
%		\gll мин өй-GA бар-IргA кирәк \\
%		I home-\gmk{Dat} go-\gmk{Inf} need \\
%		\glt \eng{I need to go home}
%	\end{xlist}
%\end{exe}

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\a
\begingl
\gla Мен үйге баруым керек. //
\glb мен үй-GA бар-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\a
\begingl
\gla Мин өйгә барырга кирәк. //
\glb мин өй-GA бар-IргA кирәк //
\glb I home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

@@

% FIXME: the following sh... stuff needs to be rephrased
The bilingual lexicon currently contains 9,269 stem to stem correspondences and was build mostly by hand. Part of toponyms and other proper names was translated semi-automatically looking up in Wikipedia links (also some of the Russian loanwords were the result of intersection of Russian and Kazakh wordlists - "автомобиль" etc).

>>>>

Without regard to the semantic alignment of these forms, the morphotactics of the cognate Kazakh distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are different.  Specifically, in both languages, the person agreement takes the form of a person copula suffix, although in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``apparently you went''), whereas in Tatar this suffix intervenes between the two pieces of the ``compound'' tense morpheme (e.g., баргансың икән ``I guess you went'').
% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.

@@

%\begin{exe}
%	\setstretch{0.8}
%	\ex
%	\begin{xlist}
%		\ex Мен үйге қайтуым керек
%		\gll мен үй-GA қайт-у-Iм керек \\
%		I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need \\
%		\glt \eng{I need to go home}
%
%		\ex Миңа өйгә кайтырга кирәк
%		\gll мин-GA өй-GA кайт-IргA кирәк \\
%		I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need \\
%		\glt \eng{I need to go home}
%	\end{xlist}
%\end{exe}

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\a
\begingl
\gla Мен үйге қайтуым керек. //
\glb мен үй-GA қайт-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\a
\begingl
\gla Миңа өйгә кайтырга кирәк. //
\glb мин-GA өй-GA кайт-IргA кирәк //
\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

@@

% FIXME: the following sh... stuff needs to be rephrased
The bilingual lexicon currently contains 9,269 stem to stem correspondences and was build mostly by hand (translating Kazakh stems unrecognized by the morphological analyzer into Tatar). Part of toponyms and other proper names was translated semi-automatically looking up in Wikipedia links (also some of the Russian loanwords were the result of intersection of Russian and Kazakh wordlists - such as "автомобиль", "гонорар" etc).
